---
title: "Data Preparation"
execute:
  warning: false
  message: false
---

# Data Preparation

This page documents how the raw challenge file becomes the clean tables the Shiny application runs on. All of it happens in one script, `data_prep.R`, which lives in the app folder. We run it once. It reads the raw JSON, builds the tables, and saves them as `.rds` files. The application then loads those files at startup instead of parsing the JSON every time, which keeps it fast.

# Setup

## Loading packages

We use the following packages to prepare the data.

- [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) to parse the JSON
- [tidyverse](https://www.tidyverse.org/) for data manipulation
- [knitr](https://cran.r-project.org/web/packages/knitr/index.html) for table displays

```{r}
#| label: load-packages
library(jsonlite)
library(tidyverse)
library(knitr)
```

## The raw data

The source is a single file, `MC1_final_00.json`, from VAST Challenge 2026 Mini-Challenge 1. It holds 23 rounds across two phases. Rounds 1 to 13 are daily snapshots from 17 May to 4 June. Rounds 14 to 23 are hourly snapshots across a single day, 5 June, from 9 AM to 6 PM. The switch from daily to hourly matters later, because it means raw message counts cannot be compared between the two phases without normalising to a per-round rate.

Each round carries three blocks:

1. `communications`, every message sent that round, with the sender, channel, recipients, message type, what it was responding to, the full text, and a private internal-state field.
2. `participants`, the agents present that round and their declared action.
3. `environment_context`, the surrounding world, including the market snapshot, news, alerts, and external actor activity.

The code below reads the file. We use `simplifyVector = FALSE` so the nested structure stays intact while we walk through it.

```{r}
#| label: read-json
#| eval: false
raw    <- fromJSON("../../data/MC1_final_00.json", simplifyVector = FALSE)
rounds <- raw$rounds
```

# Flattening the communications

The JSON is deeply nested, so the first job is to pull every message into one flat table with a single row per message. We walk the structure round by round, lifting the sender details, the channel, the content, and the three internal-state fields onto each row.

```{r}
#| label: flatten-code
#| eval: false
comms <- map(seq_along(rounds), function(ri) {
  rd <- rounds[[ri]]
  map(rd$communications, function(m) {
    ist <- m$internal_state %or% list()
    tibble(
      message_id   = m$message_id,
      round_index  = ri - 1L,
      agent_id     = m$agent_id,
      agent_label  = m$agent_label,
      channel      = m$channel,
      message_type = m$message_type,
      responding_to= m$responding_to,
      content      = m$content,
      timestamp    = m$timestamp,
      reacting     = ist$reacting,
      deliberating = ist$deliberating,
      rationalizing= ist$rationalizing
    )
  }) %>% bind_rows()
}) %>% bind_rows()
```

This produces `comms`, the table everything else is built from. Loading the saved result, we can confirm its shape.

```{r}
#| label: comms-glimpse
comms <- readRDS("../../shiny_app/my_shiny_app/data/comms.rds")
glimpse(comms)
```

There are **912 messages** across **7 agents** and **23 rounds**.

# Splitting internal from public

The central question in this challenge is when and why information crossed from inside the company to the outside. To make that measurable we tag every channel as either Internal or Public. There are six channels. Three are public, the personal post, official post, and anonymous post. The other three are internal, the comms huddle, the one-on-one chat, and the side huddle.

```{r}
#| label: channel-code
#| eval: false
public_channels <- c("anonymous_post", "official_post", "personal_post")
comms$channel_type <- ifelse(comms$channel %in% public_channels,
                             "Public", "Internal")
```

This single derived field is the backbone of the rest of the work. The table below confirms the split.

```{r}
#| label: channel-table
comms %>%
  count(channel_type, channel, name = "messages") %>%
  arrange(channel_type, desc(messages)) %>%
  kable()
```

::: callout-note
## Internal state is missing-not-random {.unnumbered}

Each message can carry a private internal state with three parts, reacting, deliberating, and rationalizing. These are filled in on only 86 of the 912 messages, so we treat them as a sparse source rather than a complete record. No table in the application aggregates over internal state. Where the fields are used, they are read as individual quotes, not as a population.
:::

# Cleaning the market data

The market snapshot gives a stock price each round, and this one needs care. The baseline prices fall steadily from 38.70 down to 31.50, which is sensible given the exposé pressure building through the buildup. On the crisis day the field breaks down. Several rounds have no price at all, including the decisive 4 PM to 6 PM window, and one round records a stock price of 180, which is the distressed-sale valuation from the news narrative bleeding into the price field.

```{r}
#| label: price-code
#| eval: false
parse_price <- function(x) {
  v <- suppressWarnings(as.numeric(str_remove_all(x, "[$,]")))
  ifelse(!is.na(v) & v >= 10 & v <= 60, v, NA_real_)
}
```

::: callout-note
## Interpretive decision: how we handled the broken price field {.unnumbered}

Charting the raw price would put a misleading line on exactly the day that matters most. Rather than trust the field, we clipped the price to a plausible band and dropped anything outside it. For the day's overall shape we lean on the sentiment state instead, which stays consistent across the whole period and moves from neutral through cautious, negative, and critical before recovering. No price-reaction or manipulation analysis is attempted, because the field cannot support one.
:::

The per-round summary table holds the cleaned price alongside the message counts and sentiment that drive the timeline view.

```{r}
#| label: round-summary-table
round_summary <- readRDS("../../shiny_app/my_shiny_app/data/round_summary.rds")
round_summary %>%
  select(round_index, n_msgs, n_public, sentiment_state, stock_price) %>%
  kable()
```

# Building the analytical tables

From the flat table we build a small set of focused tables, each one feeding a specific part of the application.

- **Round summary** gives the per-round message counts, channel split, sentiment, and cleaned price. This drives the timeline view.
- **Baseline stats** compares each agent before and during the crisis on messages per round, public share, and anonymous posts. Because the two phases have a different number of rounds, we divide by 13 for the baseline and 10 for the crisis so the rates are comparable.
- **Z-scores** measure how far each agent's volume in a given round sits from their own baseline average.
- **Network edges** resolve each reply back to the message it answered, then to that message's author, giving a who-replied-to-whom table.
- **Intent chain** is a fixed set of six messages selected by their message id, so the sequence is exact and cannot break if the ordering of the file changes.
- **Anonymous posts** pulls all 12 anonymous messages, keeping the true author stored separately so the application can hide it until the user chooses to reveal it.
- **Event pins** are the timeline markers, anchored to real hours on the crisis day.

The baseline-versus-crisis table is the clearest example. It shows how each agent's behaviour shifts between the two phases.

```{r}
#| label: baseline-stats-table
baseline_stats <- readRDS("../../shiny_app/my_shiny_app/data/baseline_stats.rds")
baseline_stats %>%
  mutate(msgs_per_round = round(msgs_per_round, 1),
         public_share = round(public_share, 2)) %>%
  select(agent_id, phase, msgs_per_round, public_share, n_anon) %>%
  arrange(agent_id, phase) %>%
  kable()
```

::: callout-note
## Interpretive decision: selecting the intent chain by id {.unnumbered}

The intent chain is a curated set of six messages, not a computed result. We selected them by message id rather than by text matching, so the chain is exact and does not silently change if the file is reordered. The selection is an analytical choice about which messages form the planning sequence, and the verdict view treats it as such.
:::

# Output

The script saves 11 `.rds` files. The key counts are below, confirming the build came out as expected.

```{r}
#| label: output-summary
tibble(
  Item = c("Messages", "Agents", "Rounds", "Messages with internal state",
           "Public messages", "Anonymous posts (all Legal)",
           "Intent chain messages"),
  Count = c(
    nrow(comms),
    n_distinct(comms$agent_id),
    n_distinct(comms$round_index),
    sum(!(is.na(comms$reacting) & is.na(comms$deliberating) & is.na(comms$rationalizing))),
    sum(comms$channel_type == "Public"),
    sum(comms$channel == "anonymous_post"),
    6
  )
) %>% kable()
```

With these files in place, the application reads clean tables and never touches the raw JSON again. The next page explores what the data looks like once it is prepared.
