---
title: "Data Preparation"
execute:
  warning: false
  message: false
---

# Data Preparation

This page documents how the raw challenge file becomes the clean tables the Shiny application runs on. All of it happens in one script, `data_prep.R`, which lives in the app folder. We run it once. It reads the raw JSON, builds the tables, and saves them as `.rds` files. The application then loads those files at startup instead of parsing the JSON every time, which keeps it fast.

# Setup

## Loading packages

We use the following packages to prepare the data.

- [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) to parse the JSON
- [tidyverse](https://www.tidyverse.org/) for data manipulation
- [knitr](https://cran.r-project.org/web/packages/knitr/index.html) for table displays

```{r}
#| label: load-packages
library(jsonlite)
library(tidyverse)
library(knitr)
```

## The raw data

The source is a single file, `MC1_final_00.json`, from VAST Challenge 2026 Mini-Challenge 1. It holds 23 rounds across two phases. Rounds 1 to 13 are baseline rounds spanning 17 May to 4 June. Rounds 14 to 23 are hourly snapshots across a single day, 5 June, from 9 AM to 6 PM. The switch from daily to hourly matters later. Because the two phases use different time resolutions, we report average messages per simulation round rather than a time-based rate, and read the comparisons as direction and magnitude of behavioural change within the challenge structure, not as precise hourly or daily ratios.

Each round carries three blocks:

1. `communications`, every message sent that round, with the sender, channel, recipients, message type, what it was responding to, the full text, and a private internal-state field.
2. `participants`, the agents present that round and their declared action.
3. `environment_context`, the surrounding world, including the market snapshot, news, alerts, and external actor activity.

The code below reads the file. We use `simplifyVector = FALSE` so the nested structure stays intact while we walk through it.

```{r}
#| label: read-json
#| eval: false
raw    <- fromJSON("../../data/MC1_final_00.json", simplifyVector = FALSE)
rounds <- raw$rounds
```

# Flattening the communications

The JSON is deeply nested, so the first job is to pull every message into one flat table with a single row per message. We walk the structure round by round, lifting the sender details, the channel, the content, and the three internal-state fields onto each row.

```{r}
#| label: flatten-code
#| eval: false
`%||%` <- function(x, y) if (is.null(x)) y else x

comms <- map(seq_along(rounds), function(ri) {
  rd <- rounds[[ri]]
  map(rd$communications, function(m) {
    ist <- m$internal_state %||% list()
    tibble(
      message_id   = m$message_id,
      round_index  = ri - 1L,
      agent_id     = m$agent_id,
      agent_label  = m$agent_label,
      channel      = m$channel,
      message_type = m$message_type,
      responding_to= m$responding_to,
      content      = m$content,
      timestamp    = m$timestamp,
      reacting     = ist$reacting,
      deliberating = ist$deliberating,
      rationalizing= ist$rationalizing
    )
  }) %>% bind_rows()
}) %>% bind_rows()
```

This produces `comms`, the table everything else is built from. Loading the saved result, we can confirm its shape.

```{r}
#| label: comms-glimpse
comms <- readRDS("../../shiny_app/my_shiny_app/data/comms.rds")
glimpse(comms)
```

There are **912 messages** across **7 agents** and **23 rounds**.

# Splitting internal from public

The central question in this challenge is when and why information crossed from inside the company to the outside. To make that measurable we tag every channel as either Internal or Public. There are six channels. Three are public, the personal post, official post, and anonymous post. The other three are internal, the comms huddle, the one-on-one chat, and the side huddle.

```{r}
#| label: channel-code
#| eval: false
public_channels <- c("anonymous_post", "official_post", "personal_post")
comms$channel_type <- ifelse(comms$channel %in% public_channels,
                             "Public", "Internal")
```

This single derived field is the backbone of the rest of the work. The table below confirms the split.

```{r}
#| label: channel-table
comms %>%
  count(channel_type, channel, name = "messages") %>%
  arrange(channel_type, desc(messages)) %>%
  kable()
```

::: callout-note
## Internal state is sparse and selectively populated {.unnumbered}

Each message can carry a private internal state with three parts: reacting, deliberating, and rationalizing. These fields are populated for only 86 of the 912 messages, so we treat them as a sparse source rather than a complete record. No table in the application aggregates the internal-state fields. Where they are used, they are presented as individual quotations rather than interpreted as representative of the full dataset.
:::

# Handling the market data

The market snapshot gives a stock price each round, and this one needs care. The baseline prices fall steadily from 38.70 down to 31.50, which is sensible given the exposé pressure building through the buildup. On the crisis day the field breaks down. Several rounds have no price at all, including the decisive 4 PM to 6 PM window, and one round records a stock price of 180, which is the distressed-sale valuation from the news narrative bleeding into the price field.

```{r}
#| label: price-code
#| eval: false
parse_price <- function(x) {
  suppressWarnings(as.numeric(str_remove_all(x, "[$,]")))
}
# the crisis-day field is contaminated: several rounds are missing and one
# records 180, the distressed-sale valuation bleeding in from the news. We
# parse the raw value, flag the 180 reading, and exclude the field from the
# verdict because of missingness and contamination.
```

::: callout-note
## Interpretive decision: how we handled the unreliable price field {.unnumbered}

Charting the raw price would produce a misleading series on exactly the day that matters most. Rather than impose an analyst-selected valid range, we retain the parsed values, flag the contextually inconsistent reading of 180, and exclude the stock-price field from substantive analysis because of its missingness and apparent contamination. For contextual orientation, we use the sentiment state, which remains populated across the full period and moves from neutral through cautious, negative, and critical before recovering. The project therefore makes no claims about price reactions or market manipulation.
:::

The per-round summary table retains the parsed price field alongside the message counts and sentiment used by the timeline view.

```{r}
#| label: round-summary-table
round_summary <- readRDS("../../shiny_app/my_shiny_app/data/round_summary.rds")
round_summary %>%
  select(round_index, n_msgs, n_public, sentiment_state, stock_price) %>%
  kable()
```

# Building the analytical tables

From the flat table we build a small set of focused tables, each one feeding a specific part of the application.

- **Round summary** gives the per-round message counts, channel split, sentiment, and parsed price. This drives the timeline view.
- **Baseline stats** compares each agent before and during the crisis on messages per simulation round, public share, and anonymous posts. We divide by 13 for the baseline and 10 for the crisis to get an average per round. Because the baseline rounds are daily and the crisis rounds hourly, this is a per-round average within the challenge structure, not a comparable time-based rate, and is read as direction and magnitude rather than an exact multiple.
- **Z-scores** measure how many baseline standard deviations each agent's message count in a given simulation round lies above or below that agent's baseline mean.
- **Network edges** resolve each reply back to the message it answered, then to that message's author, giving a who-replied-to-whom table.
- **Intent chain** is a curated set of six messages selected by message id and arranged in a fixed sequence, so the same messages appear in the intended order even if the source file is reordered.
- **Anonymous posts** retains the twelve anonymous-channel messages and their recorded author field, so the application can control when authorship is displayed.
- **Event pins** are the timeline markers, anchored to real hours on the crisis day.

The baseline-versus-crisis table is the clearest example. It shows how each agent's behaviour shifts between the two phases.

```{r}
#| label: baseline-stats-table
baseline_stats <- readRDS("../../shiny_app/my_shiny_app/data/baseline_stats.rds")
baseline_stats %>%
  mutate(msgs_per_round = round(msgs_per_round, 1),
         public_share = round(public_share, 2)) %>%
  select(agent_id, phase, msgs_per_round, public_share, n_anon) %>%
  arrange(agent_id, phase) %>%
  kable()
```

::: callout-note
## Interpretive decision: selecting the intent chain by id {.unnumbered}

The intent chain is a curated set of six messages, not a computed result. We selected them by message id rather than by text matching, and arranged them in a fixed sequence, so the chain is exact and does not silently change if the file is reordered. The selection is an analytical choice about which messages form the planning sequence, and the verdict view treats it as such.
:::

# Output

The script saves ten core .rds files and, when tidytext is available, one optional TF-IDF file. The key counts are below

```{r}
#| label: output-summary
tibble(
  Item = c("Messages", "Agents", "Rounds", "Messages with internal state",
           "Public messages", "Anonymous posts",
           "Intent chain messages"),
  Count = c(
    nrow(comms),
    n_distinct(comms$agent_id),
    n_distinct(comms$round_index),
    sum(!(is.na(comms$reacting) & is.na(comms$deliberating) & is.na(comms$rationalizing))),
    sum(comms$channel_type == "Public"),
    sum(comms$channel == "anonymous_post"),
    6
  )
) %>% kable()
```

With these files in place, the application reads clean tables and never touches the raw JSON again. The next page explores what the data looks like once it is prepared.
